---
title: "Delayed Monetary Losses: Reproducible Analysis in R"
subtitle: "Adjusting-Amount and Delayed Losses Questionnaire measures"
author: "Haoran Wan"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    code-summary: "Show analysis code"
    self-contained: true
    theme: default
execute:
  echo: true
  warning: false
  message: false
engine: knitr
---

## Scope

This document reproduces the primary measurement-validation analyses reported in:

> Wan, H., Green, L., & Myerson, J. (2024). Delayed monetary losses: Do different procedures and discounting measures assess the same construct? *Behavioural Processes, 222*, 105101. <https://doi.org/10.1016/j.beproc.2024.105101>

The analysis reproduces Figure 1 and Tables 2-4: group-level model fits, convergent validity between procedures, concurrent validity between scoring measures, alternate-forms reliability across amounts, and consistency of individual discounting patterns. The final section checks the computed results against the values printed in the article.

The anonymized data are archived on the [Open Science Framework](https://osf.io/emb2q/). The repository copy is stored in `data/Data_AdjAmt_DLQ.csv` for computational reproducibility.

## Setup

The workflow intentionally uses base R plus `knitr`, which minimizes dependencies and makes the analysis straightforward to audit.

```{r setup}
options(digits = 4, scipen = 5, width = 100)
set.seed(20240913)

locate_data <- function() {
  candidates <- c(
    "data/Data_AdjAmt_DLQ.csv",
    "../data/Data_AdjAmt_DLQ.csv"
  )
  existing <- candidates[file.exists(candidates)]
  if (length(existing) == 0) {
    stop(
      "data/Data_AdjAmt_DLQ.csv was not found. Run from the repository root ",
      "or the Analysis directory."
    )
  }
  normalizePath(existing[[1]])
}

data_path <- locate_data()
amount_order_adj <- c("$90", "$240", "$1,500")
amount_order_dlq <- c("$90", "$165", "$240")
common_amounts <- c("$90", "$240")
pattern_order <- c("Typical", "Always immediate", "Atypical", "Inconsistent")
```

### Analysis functions

```{r functions}
auc <- function(delay, subjective_value) {
  order_index <- order(delay)
  x <- delay[order_index] / 108
  y <- subjective_value[order_index]
  sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2)
}

simple_hyperbola <- function(delay, log_k) {
  1 / (1 + exp(log_k) * delay)
}

hyperboloid <- function(delay, log_k, s) {
  1 / (1 + exp(log_k) * delay)^s
}

logistic_growth <- function(log_k, midpoint, slope) {
  1 / (1 + exp(-(log_k - midpoint) * slope))
}

r_squared <- function(observed, predicted) {
  1 - sum((observed - predicted)^2) / sum((observed - mean(observed))^2)
}

score_adj_amt <- function(group) {
  group <- group[order(group$Delay), ]
  objective <- function(log_k) {
    sum((group$RSV - simple_hyperbola(group$Delay, log_k))^2)
  }
  log_k <- optimize(objective, interval = c(-30, 10), tol = 1e-12)$minimum
  data.frame(
    ID = group$ID[[1]],
    Amount = group$Amount[[1]],
    auc_adj_amt = auc(group$Delay, group$RSV),
    logk_adj_amt = log_k
  )
}

canonical_dlq_k <- function(k) {
  ifelse(
    abs(k - 0.00088) < 1e-12, 0.00089,
    ifelse(
      abs(k - 0.0018) < 1e-12, 0.0023,
      ifelse(abs(k - 0.0058) < 1e-12 | abs(k - 0.0059) < 1e-12, 0.0055, k)
    )
  )
}

score_dlq <- function(group) {
  choices <- group$Choice
  raw_k <- group$k_raw
  scoring_k <- canonical_dlq_k(raw_k)
  candidates <- sort(unique(scoring_k))
  consistency <- vapply(
    candidates,
    function(candidate) {
      sum(
        (choices == 0 & scoring_k <= candidate) |
          (choices == 1 & scoring_k >= candidate)
      )
    },
    numeric(1)
  )
  best_candidates <- candidates[consistency == max(consistency)]
  best_raw_k <- raw_k[scoring_k %in% best_candidates]

  data.frame(
    ID = group$ID[[1]],
    Amount = group$Amount[[1]],
    prop_imm_dlq = mean(choices),
    logk_dlq = mean(log(best_raw_k))
  )
}

correlation_row <- function(x, y) {
  complete <- complete.cases(x, y)
  test <- cor.test(x[complete], y[complete], method = "pearson")
  c(r = unname(test$estimate), p = test$p.value, n = sum(complete))
}

format_p <- function(p) {
  ifelse(p < 0.001, "< .001", sprintf("%.3f", p))
}

collapse_patterns <- function(values) {
  if (length(unique(values)) == 1) values[[1]] else "Inconsistent"
}
```

## Data and quality checks

The public file contains the analyzed Prolific sample after the exclusions reported in the article. `Choice = 1` denotes selection of the smaller, immediate payment on the DLQ.

```{r data}
raw <- read.csv(data_path, stringsAsFactors = FALSE, check.names = FALSE)
raw <- raw[, -1]

adj_amt <- subset(raw, Procedure == "Adj-Amt")
names(adj_amt)[names(adj_amt) == "IV"] <- "Delay"
names(adj_amt)[names(adj_amt) == "DV"] <- "RSV"

dlq <- subset(raw, Procedure == "MCQ")
names(dlq)[names(dlq) == "IV"] <- "k_raw"
names(dlq)[names(dlq) == "DV"] <- "Choice"
dlq$Procedure <- "DLQ"
dlq$Amount <- unname(c(
  "$75-$105" = "$90",
  "$150-$180" = "$165",
  "$225-$255" = "$240"
)[dlq$Amount])

stopifnot(
  length(unique(raw$ID)) == 431,
  nrow(adj_amt) == 5172,
  nrow(dlq) == 11637,
  identical(sort(unique(adj_amt$Delay)), c(4, 18, 60, 108)),
  identical(sort(unique(dlq$Choice)), c(0, 1)),
  all(table(adj_amt$ID, adj_amt$Amount) == 4),
  all(table(dlq$ID, dlq$Amount) == 9)
)

data.frame(
  procedure = c("Adjusting-Amount", "DLQ"),
  participants = c(length(unique(adj_amt$ID)), length(unique(dlq$ID))),
  observations = c(nrow(adj_amt), nrow(dlq)),
  amount_conditions = c(3, 3),
  observations_per_participant_amount = c(4, 9)
)
```

## Individual discounting scores

For the Adj-Amt procedure, the atheoretical measure is area under the curve (AuC), and the theoretical measure is the logarithm of the best-fitting *k* from a simple hyperbola. For the DLQ, the corresponding measures are immediate-choice proportion and the logarithm of the *k* value most consistent with the participant's choices.

For delayed losses, lower AuC and lower immediate-choice proportion indicate steeper discounting, whereas lower log(*k*) indicates shallower discounting. Consequently, concurrent correlations between atheoretical and theoretical measures are expected to be negative.

```{r scores}
adj_groups <- split(adj_amt, interaction(adj_amt$ID, adj_amt$Amount, drop = TRUE))
adj_scores <- do.call(rbind, lapply(adj_groups, score_adj_amt))
row.names(adj_scores) <- NULL

dlq_groups <- split(dlq, interaction(dlq$ID, dlq$Amount, drop = TRUE))
dlq_scores <- do.call(rbind, lapply(dlq_groups, score_dlq))
row.names(dlq_scores) <- NULL

stopifnot(nrow(adj_scores) == 431 * 3, nrow(dlq_scores) == 431 * 3)
```

## Group-level discounting (Figure 1)

```{r group-models, fig.width=10, fig.height=4.8, fig.cap="Group-level replication of Figure 1. Points are observed means; lines are the fitted hyperboloid (Adj-Amt) and logistic growth (DLQ) functions."}
adj_group <- aggregate(RSV ~ Amount + Delay, adj_amt, mean)
dlq_group <- aggregate(Choice ~ Amount + k_raw, dlq, mean)

colors_adj <- c("$90" = "#0072B2", "$240" = "#E69F00", "$1,500" = "#009E73")
symbols_adj <- c("$90" = 21, "$240" = 22, "$1,500" = 24)
line_types_adj <- c("$90" = 1, "$240" = 2, "$1,500" = 3)

adj_fits <- list()
adj_r2 <- numeric(length(amount_order_adj))
names(adj_r2) <- amount_order_adj
for (amount in amount_order_adj) {
  group <- subset(adj_group, Amount == amount)
  objective <- function(parameters) {
    sum((group$RSV - hyperboloid(group$Delay, parameters[[1]], parameters[[2]]))^2)
  }
  fit <- optim(c(-4, 1), objective, method = "Nelder-Mead", control = list(reltol = 1e-12))
  adj_fits[[amount]] <- fit$par
  adj_r2[[amount]] <- r_squared(
    group$RSV,
    hyperboloid(group$Delay, fit$par[[1]], fit$par[[2]])
  )
}

dlq_colors <- c("$90" = "#0072B2", "$165" = "#E69F00", "$240" = "#009E73")
dlq_symbols <- c("$90" = 21, "$165" = 25, "$240" = 24)
dlq_line_types <- c("$90" = 1, "$165" = 2, "$240" = 3)
dlq_labels <- c("$90" = "$75-105", "$165" = "$150-180", "$240" = "$225-255")

dlq_fits <- list()
dlq_r2 <- numeric(length(amount_order_dlq))
names(dlq_r2) <- amount_order_dlq
for (amount in amount_order_dlq) {
  group <- subset(dlq_group, Amount == amount)
  log_k <- log(group$k_raw)
  objective <- function(parameters) {
    sum((group$Choice - logistic_growth(log_k, parameters[[1]], parameters[[2]]))^2)
  }
  fit <- optim(c(-10, 1), objective, method = "Nelder-Mead", control = list(reltol = 1e-12))
  dlq_fits[[amount]] <- fit$par
  dlq_r2[[amount]] <- r_squared(
    group$Choice,
    logistic_growth(log_k, fit$par[[1]], fit$par[[2]])
  )
}

old_par <- par(mfrow = c(1, 2), mar = c(4.3, 4.5, 2.5, 1), las = 1)
plot(
  NA, xlim = c(0, 113), ylim = c(0, 1.03), xaxt = "n",
  xlab = "Delay (months)", ylab = "Relative subjective value",
  main = "Adjusting-Amount"
)
axis(1, at = c(4, 18, 60, 108))
for (amount in amount_order_adj) {
  group <- subset(adj_group, Amount == amount)
  points(
    group$Delay, group$RSV, pch = symbols_adj[[amount]], bg = colors_adj[[amount]],
    col = "black", cex = 1.3
  )
  x_grid <- seq(0, 108, length.out = 300)
  parameters <- adj_fits[[amount]]
  lines(
    x_grid, hyperboloid(x_grid, parameters[[1]], parameters[[2]]),
    col = colors_adj[[amount]], lty = line_types_adj[[amount]], lwd = 2
  )
}
legend(
  "bottomright", legend = amount_order_adj, pch = symbols_adj, pt.bg = colors_adj,
  col = colors_adj, lty = line_types_adj, bty = "n", title = "Amount"
)

plot(
  NA, xlim = range(log(dlq_group$k_raw)), ylim = c(0, 1.03), xaxt = "n",
  xlab = expression(paste("Question ", italic(k), " parameter")),
  ylab = "Proportion choosing immediate payment",
  main = "Delayed Losses Questionnaire"
)
axis(1, at = log(c(0.00001, 0.0001, 0.001, 0.01)), labels = c("0.00001", "0.0001", "0.001", "0.01"))
for (amount in amount_order_dlq) {
  group <- subset(dlq_group, Amount == amount)
  points(
    log(group$k_raw), group$Choice, pch = dlq_symbols[[amount]],
    bg = dlq_colors[[amount]], col = "black", cex = 1.3
  )
  x_grid <- seq(min(log(group$k_raw)), max(log(group$k_raw)), length.out = 300)
  parameters <- dlq_fits[[amount]]
  lines(
    x_grid, logistic_growth(x_grid, parameters[[1]], parameters[[2]]),
    col = dlq_colors[[amount]], lty = dlq_line_types[[amount]], lwd = 2
  )
}
legend(
  "bottomright", legend = dlq_labels, pch = dlq_symbols, pt.bg = dlq_colors,
  col = dlq_colors, lty = dlq_line_types, bty = "n", title = "Amount"
)
par(old_par)

model_fit <- rbind(
  data.frame(procedure = "Adj-Amt", amount = names(adj_r2), R2 = unname(adj_r2)),
  data.frame(procedure = "DLQ", amount = names(dlq_r2), R2 = unname(dlq_r2))
)
knitr::kable(model_fit, digits = 3)
```

All group-level fits exceed the thresholds reported in the article: R² > .95 for Adj-Amt and R² > .97 for the DLQ.

## Convergent validity between procedures (Table 2)

Only the two amounts common to both procedures ($90 and $240) enter the primary convergent-validity analysis.

```{r convergent-validity}
common_scores <- merge(adj_scores, dlq_scores, by = c("ID", "Amount"))
common_scores <- subset(common_scores, Amount %in% common_amounts)

table2_rows <- list()
row_index <- 1
for (amount in common_amounts) {
  group <- subset(common_scores, Amount == amount)
  for (measure in c("AuC / immediate-choice proportion", "log(k)")) {
    values <- if (measure == "log(k)") {
      correlation_row(group$logk_adj_amt, group$logk_dlq)
    } else {
      correlation_row(group$auc_adj_amt, group$prop_imm_dlq)
    }
    table2_rows[[row_index]] <- data.frame(
      amount = amount,
      measure = measure,
      r = values[["r"]],
      p = values[["p"]],
      n = values[["n"]]
    )
    row_index <- row_index + 1
  }
}
table2 <- do.call(rbind, table2_rows)
table2_display <- transform(table2, p = format_p(p))
knitr::kable(table2_display, digits = 3)
```

## Concurrent validity and alternate-forms reliability (Table 3)

```{r concurrent-validity}
concurrent_rows <- list()
row_index <- 1
for (amount in amount_order_adj) {
  group <- subset(adj_scores, Amount == amount)
  values <- correlation_row(group$auc_adj_amt, group$logk_adj_amt)
  concurrent_rows[[row_index]] <- data.frame(
    procedure = "Adj-Amt", amount = amount,
    r = values[["r"]], p = values[["p"]], n = values[["n"]]
  )
  row_index <- row_index + 1
}
for (amount in amount_order_dlq) {
  group <- subset(dlq_scores, Amount == amount)
  values <- correlation_row(group$prop_imm_dlq, group$logk_dlq)
  concurrent_rows[[row_index]] <- data.frame(
    procedure = "DLQ", amount = amount,
    r = values[["r"]], p = values[["p"]], n = values[["n"]]
  )
  row_index <- row_index + 1
}
table3a <- do.call(rbind, concurrent_rows)
table3a_display <- transform(table3a, p = format_p(p))
knitr::kable(table3a_display, digits = 3)
```

```{r reliability}
reliability_matrix <- function(scores, value, amount_order) {
  wide <- reshape(
    scores[c("ID", "Amount", value)],
    idvar = "ID", timevar = "Amount", direction = "wide"
  )
  columns <- paste(value, amount_order, sep = ".")
  result <- cor(wide[columns], use = "pairwise.complete.obs")
  dimnames(result) <- list(amount_order, amount_order)
  result
}

table3b_adj <- reliability_matrix(adj_scores, "auc_adj_amt", amount_order_adj)
table3b_dlq <- reliability_matrix(dlq_scores, "prop_imm_dlq", amount_order_dlq)
table3c_adj <- reliability_matrix(adj_scores, "logk_adj_amt", amount_order_adj)
table3c_dlq <- reliability_matrix(dlq_scores, "logk_dlq", amount_order_dlq)

knitr::kable(table3b_adj, digits = 3, caption = "Adj-Amt AuC")
knitr::kable(table3b_dlq, digits = 3, caption = "DLQ immediate-choice proportion")
knitr::kable(table3c_adj, digits = 3, caption = "Adj-Amt log(k)")
knitr::kable(table3c_dlq, digits = 3, caption = "DLQ log(k)")
```

## Individual discounting patterns (Table 4)

Patterns are classified separately for the $90 and $240 conditions, then combined within each procedure. "Inconsistent" means that a participant received different classifications for the two amounts.

```{r patterns}
classify_adj_group <- function(group) {
  correlation_value <- suppressWarnings(cor(group$RSV, log(group$Delay)))
  if (is.na(correlation_value)) correlation_value <- 0
  mean_value <- mean(group$RSV)
  if (correlation_value < 0 || isTRUE(all.equal(mean_value, min(adj_amt$RSV)))) {
    "Typical"
  } else if (correlation_value > 0) {
    "Atypical"
  } else if (isTRUE(all.equal(mean_value, max(adj_amt$RSV)))) {
    "Always immediate"
  } else {
    "Typical"
  }
}

classify_dlq_group <- function(group) {
  # Pattern classification uses each item's original k, as in the article.
  correlation_value <- suppressWarnings(cor(group$Choice, log(group$k_raw)))
  if (is.na(correlation_value)) correlation_value <- 0
  mean_choice <- mean(group$Choice)
  if (correlation_value > 0 || isTRUE(all.equal(mean_choice, 0))) {
    "Typical"
  } else if (correlation_value < 0) {
    "Atypical"
  } else if (isTRUE(all.equal(mean_choice, 1))) {
    "Always immediate"
  } else {
    "Unclassified"
  }
}

adj_pattern_data <- subset(adj_amt, Amount %in% common_amounts)
adj_pattern_groups <- split(
  adj_pattern_data,
  interaction(adj_pattern_data$ID, adj_pattern_data$Amount, drop = TRUE)
)
adj_amount_patterns <- do.call(
  rbind,
  lapply(adj_pattern_groups, function(group) {
    data.frame(ID = group$ID[[1]], Amount = group$Amount[[1]], pattern = classify_adj_group(group))
  })
)

dlq_pattern_data <- subset(dlq, Amount %in% common_amounts)
dlq_pattern_groups <- split(
  dlq_pattern_data,
  interaction(dlq_pattern_data$ID, dlq_pattern_data$Amount, drop = TRUE)
)
dlq_amount_patterns <- do.call(
  rbind,
  lapply(dlq_pattern_groups, function(group) {
    data.frame(ID = group$ID[[1]], Amount = group$Amount[[1]], pattern = classify_dlq_group(group))
  })
)

adj_patterns <- aggregate(pattern ~ ID, adj_amount_patterns, collapse_patterns)
names(adj_patterns)[[2]] <- "Adj_Amt"
dlq_patterns <- aggregate(pattern ~ ID, dlq_amount_patterns, collapse_patterns)
names(dlq_patterns)[[2]] <- "DLQ"
patterns <- merge(adj_patterns, dlq_patterns, by = "ID")

table4 <- table(
  `Adj-Amt` = factor(patterns$Adj_Amt, levels = pattern_order),
  DLQ = factor(patterns$DLQ, levels = pattern_order)
)
knitr::kable(table4, caption = "Cross-tabulation of individual discounting patterns")
```

```{r consistency-test}
expected <- outer(rowSums(table4), colSums(table4)) / sum(table4)
# The article's test counts matching Typical and Always-immediate patterns.
# Atypical and Inconsistent classifications are not counted as matches.
observed_same <- table4["Typical", "Typical"] +
  table4["Always immediate", "Always immediate"]
expected_same <- expected["Typical", "Typical"] +
  expected["Always immediate", "Always immediate"]
published_chi_square <- (observed_same - expected_same)^2 / expected_same
published_p <- pchisq(published_chi_square, df = 1, lower.tail = FALSE)

data.frame(
  observed_same = observed_same,
  expected_same = expected_same,
  same_pattern_proportion = observed_same / nrow(patterns),
  chi_square = published_chi_square,
  df = 1,
  p = format_p(published_p)
) |>
  knitr::kable(digits = 3, caption = "Published same-pattern consistency test")
```

## Reproduction checks

These assertions fail if future code or data changes no longer reproduce the article within rounding tolerance.

```{r reproduction-checks}
published_table2 <- c(0.735, 0.728, 0.777, 0.773)
stopifnot(all(abs(table2$r - published_table2) <= 0.0006))

published_table4 <- matrix(
  c(
    255, 1, 1, 12,
    5, 56, 0, 10,
    5, 2, 0, 1,
    62, 8, 0, 13
  ),
  nrow = 4,
  byrow = TRUE
)
stopifnot(
  all(as.matrix(table4) == published_table4),
  abs(published_chi_square - 42.7) <= 0.1,
  abs(observed_same / nrow(patterns) - 0.722) <= 0.0005
)

cat("All primary results reproduce the published values within rounding tolerance.\n")
```

## Computational environment

```{r session-info}
sessionInfo()
```
